---
title: "p8105_hw2_WL3011"
author: "Weiqi Liang"
date: "`r Sys.Date()`"
output: html_document
---

## Setup file
```{r setup, message = FALSE, results='hide'}
library(tidyverse)
library(dplyr)
```

## 1. Problem 1
Firstly load the NYC subway dataset and clean the data.

**Data cleaning:**

*   Retained columns: `Line`, `Station Name`, `Station Latitude`, `Station Longitude`, `Route1`:`Route11`, `Entry`, `Vending`, `Entrance Type`, `ADA`;
*   Converted the `Entry` variable from "YES" and "NO" to logical TRUE and FALSE.
```{r}
# Load the NYC Subway csv
subway_df = 
  read.csv("./NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
           na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude,
         route1:route11, entry, vending, entrance_type, ada) |>
  mutate(
    entry_logical = case_match(
      entry, 
      "YES" ~ TRUE,
      "NO"  ~ FALSE
      )
    )
```
The dataset now is tidy, with each row representing a station entrance and all columns having consistent data types. it has `r nrow(subway_df)` rows and `r ncol(subway_df)` columns.
